Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-05-09 ms/spaceness.tex
andy & cj overleaf edits w typo fixes

Two areas in which spatial structure may be particularly important are demographic inference and genome-wide association studies (GWAS). Effective population size is a fundamental parameter of population genetics and shapes expectations of the efficacy of selection, the rate of drift, among other evolutionary forces (\textit{c.f.} \cite{gillespie2001population}). Inferring a trajectory of population sizes through time has therefore become a major goal in population genetics, with many recent methods proposed to accomplish this task \citep{Liu2015,Terhorst2016,Schiffels2014,Sheehan2013}. All of these methods assume that the populations being modeled are approximately randomly mating (though this also depends on how researchers interpret the inferred parameter $\Theta=4N\mu$). Because isolation by distance is expected to change the rate of drift in a population \citep{Maruyama1972}, it may cause biases in inferred population sizes. However the effects of continuous spatial structure on demographic inference has not to our knowledge been previously investigated.

@@

% The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1942,Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. 
The degree to which genetic relationships are geographically correlated
is determined by the interaction between rates of migration (the chance that one is descended from a distant location)
and coalescence (the chance that they share a parent).
A key early observation Wright model is that this balance is nicely summarized by the ``neighborhood size'',
defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
or the number of potential parents of a randomly chosen individual. Empirical estimates of neighborhood size vary hugely across species (see Discussion), but for reference estimates in human populations range from 40 to over 5,000 depending on the population and method of estimation (\ref{table:NStable}), and \cite{Wright1946} estimated values of 500-1000 for \textit{Drosophila pseudoobscura}.
\cite{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population 
approaches the random mating expectation when $\rho\sigma^2 > 1$, 
and proposed that this had the important implication that most population genetic expectations for randomly mating populations 
could be applied to continuously distributed populations with relatively little error.

@@

At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ where $L$ is the expected lifespan.
who disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
in both the $x$ and $y$ coordinates, reflected to stay within the species range.
Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2 / 2 \sigma^2)$.

>>>>

Two areas in which spatial structure may be particularly important are demographic inference and genome-wide association studies (GWAS). Effective population size is a fundamental parameter of population genetics and shapes expectations of the efficacy of selection and the rate of drift, among other evolutionary forces (\textit{c.f.} \cite{gillespie2001population}). Inferring a trajectory of population sizes through time has therefore become a major goal in population genetics, with many recent methods proposed to accomplish this task \citep{Liu2015,Terhorst2016,Schiffels2014,Sheehan2013}. All of these methods assume that the populations being modeled are approximately randomly mating (though this also depends on how researchers interpret the inferred parameter $\Theta=4N\mu$). Because isolation by distance is expected to change the rate of drift in a population \citep{Maruyama1972}, it may cause biases in inferred population sizes. However the effect of continuous spatial structure on demographic inference has not to our knowledge been previously investigated.

@@

% The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1942,Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. 
The degree to which genetic relationships are geographically correlated
is determined by the interaction between rates of migration (the chance that one is descended from a distant location)
and coalescence (the chance that they share a parent).
A key early observation Wright model is that this balance is nicely summarized by the ``neighborhood size'',
defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance and $\rho$ is population density. 
This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
or the number of potential parents of a randomly chosen individual. Empirical estimates of neighborhood size vary hugely across species (see Discussion), but for reference estimates in human populations range from 40 to over 5,000 depending on the population and method of estimation (Table \ref{table:NStable}).
%\cite{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population approaches the random mating expectation when $\rho\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

@@

At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ where $L$ is the expected lifespan.
Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
in both the $x$ and $y$ coordinates, reflected to stay within the species range.
Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2 / 2 \sigma^2)$.
